### 2018-03-10 paper/background.tex
Rewrite S2.1-S2.2

If the user modifies the sources of the utility library and runs \Make, it will
perform a full rebuild, because all three tasks transitively depend on the
library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
user modifies \cmd{main.c} then a partial rebuild is sufficient: indeed, the
file \cmd{util.o} does not need to be rebuilt, since its inputs have not
changed, see Fig.~\ref{fig-make}(c). Files that have changed since the previous
build are called \emph{dirty}.

The requirement to \emph{execute tasks at most once and only if they transitively
depend on dirty inputs} is essential for build systems, it is their raison
d'\^etre. We will call build systems that satisfy this requirement \emph{minimal}.

To achieve minimality \Make relies on two main ideas. First, it uses \emph{file
modification time} to detect the files that are dirty: a file is marked dirty if
it was modified since the previous build. Second, it constructs a complete task
dependency graph from the information contained in the makefile and executes
tasks in a \emph{topological order}.

@@

\Excel is a build system that can deal with dynamic dependencies albeit at the
cost of sacrificing minimality. Consider the following spreadsheet comprising
four input cells \cmd{A1-A3} and \cmd{B1}, and a single task computing the sum
of values \cmd{A1} $+\cdots+$ \cmd{An}, where \cmd{n} is specified indirectly in
\cmd{B1}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
A3 = 30
B1 = 2
C1 = SUM(A1:INDIRECT("A" & B1))
\end{minted}
\vspace{1mm}

\noindent
The output \cmd{C1} here is analogous to the \cmd{release.tar} from the previous
example: one cannot statically determine the dependencies of the associated task.
To deal with this, \Excel takes a simple conservative approach: it marks all
cells with dynamic dependencies as dirty and therefore always recomputes them
during a rebuild. This guarantees that the build result is correct, but clearly
violates the minimality requirement: if the user modifies \cmd{A3}, \Excel will
recompute \cmd{C1} even though it does not transitively depend on \cmd{A3}.

Note that \Excel's approach to marking cells as dirty is a relatively minor
variation of the approach used by \Make: a cell is considered dirty if it was
modified since the previous build \emph{or if it is volatile}, where the notion
of volatility includes dynamic dependencies, as well as \emph{non-determinism}
that will be discussed in~\S\ref{sec-non-determinism}. We therefore put both
\Make and \Excel in the same category of build systems that associate a
\emph{dirty bit} with every file or cell. (To avoid the ``file or cell''
nuisance, we introduce the abstract notion of \emph{keys}
in~\S\ref{sec-background-vocabulary}.)

\Excel's recalculation algorithm~\cite{excel_recalc} on the other hand is
significantly different from \Make. Since it is impossible to construct the full
dependency graph upfront in the presence of dynamic dependencies, \Excel
maintains the \emph{calculation chain}, which is an approximation to the correct
topological order produced by the previous build. During recalculation, \Excel
processes cells in this order, but can \emph{defer recalculation of a cell} by
moving it down the chain when one of its newly discovered dynamic dependencies
is dirty. We discuss this algorithm in more detail in~\S\ref{sec-implementations}.

@@

\begin{figure}[h]
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\vspace{-1mm}
\caption{Task dependency graph produced after the previous build.}
\vspace{3mm}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\vspace{-1mm}
\caption{Since \cmd{docs.txt} is dirty, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
\end{figure}

>>>>

If the user modifies the sources of the utility library and runs \Make, it will
perform a \emph{full rebuild}, because all three tasks transitively depend on
the library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient: indeed,
the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
\emph{acyclic} then each task needs to be executed at most once. Cyclic task
dependencies are typically not allowed in build systems although there are rare
exceptions, see~\S\ref{sec-iterative-compute}.

The following property is essential for build systems, it is their raison d'\^etre.

\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
most once per build and only if they transitively depend on files that changed
since the previous build.}
\vspace{2mm}

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification time} to detect the files that changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a
complete task dependency graph from the information contained in the makefile
and executes tasks in a \emph{topological order}. See further details
in~\S\ref{sec-implementation-make}.

\Excel is a build system in disguise. Consider the following simple spreadsheet.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
B1 = A1 + A2
\end{minted}
\vspace{1mm}

\noindent
There are two input cells \cmd{A1} and \cmd{A2}, and a single task that computes
the sum of their values, writing the result into the cell \cmd{B1}. If either of
the inputs change, \Excel will recompute the result.

Unlike \Make, \Excel does not need to know all task dependencies upfront. Some
dependencies may change \emph{dynamically} according to computation results. For
example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
B1 = IF(C1=1,A1,A2)
C1 = 1
\end{minted}
\vspace{1mm}

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and this is
not known statically. One might say that the value of \cmd{C1} is statically
known in this example, but imagine that it is a result of a long computation
chain -- its value will then only become available during the build.

\Excel handles this example correctly: if \cmd{C1=1} and the user changes
\cmd{A2}, \Excel will not recompute the result \cmd{B1}. Alas, other forms of
dynamic dependencies can force \Excel to perform unnecessary computation.
Consider the following modification of the above example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
A1 = 10
A2 = 20
B1 = INDIRECT("A" & C1)
C1 = 1
\end{minted}
\vspace{1mm}

\noindent
The new version uses the \cmd{INDIRECT} function, which allows us to reference a
cell indirectly by a text string that is not necessarily known in advance. The
current implementation of \Excel recomputes indirect references in every
build~\cite{excel_recalc}. This clearly violates the minimality property: if
\cmd{C1=1} and the user modifies \cmd{A2}, \Excel will recompute \cmd{B1},
potentially triggering further unnecessary recalculation even though \cmd{B1}
does not transitively depend on \cmd{A2}.

\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. Since it is impossible to construct the full dependency graph upfront in
the presence of dynamic dependencies, \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We further refer to this
algorithm as \emph{reordering}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

@@

@@

\begin{figure}[h]
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\vspace{-1mm}
\caption{Dependency graph produced after the previous build.}
\vspace{3mm}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\vspace{-1mm}
\caption{Since \cmd{docs.txt} is dirty, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
\end{figure}
